---
title: "Association Between individual Health Management, Income, and Insurance Coverage in Preventive Healthcare Among U.S. Adults"
format: 
  pdf: 
    keep-tex: true
    fig-width: 6
    fig-height: 4
evaluate: 
  warning: false
  message: false
author:
  - name: Sandesh Manjunath Raykar
    affiliations:
      - name: California State University, Chico
        department: Data Science and Analytics
  - name: Daniel Lee
    affiliations:
       - name: California State University, Chico
         department: Data Science and Analytics
abstract: |
 This study examines two interrelated questions using data from the 2016–2017 American Health Values Survey (AHVS), a nationally representative dataset of U.S. adults collected by NORC at the University of Chicago. The first question investigates whether individuals who report putting greater effort into managing their personal health—through behaviors such as regular exercise, healthy eating, stress reduction, and routine medical engagement—are less likely to have been diagnosed by a doctor with high blood pressure. The second question explores how annual household income and insurance coverage are associated with the time since an individual’s last routine medical checkup, reflecting differences in access to preventive healthcare services across socioeconomic groups. 
  
  
keywords: [health effort, income, checkup]
bibliography: references.bib
link-citations: true
editor_options: 
  chunk_output_type: console
---

## Introduction and Background {#sec-intro}

Preventive healthcare plays a critical role in early disease detection and reducing long-term health costs. Regular checkups, screenings, and healthy lifestyle behaviors help identify potential health risks before they become serious, ultimately improving quality of life and reducing the financial burden on individuals and the healthcare system. Despite these benefits, engagement in preventive health practices varies widely among Americans, reflecting differences in awareness, motivation, and access to care.

To better understand patterns in Americans health, this study takes two complementary perspectives. The first focuses on individual effort to maintain or improve health, capturing personal behaviors such as exercise, healthy eating, and proactive health management. The second examines access to routine medical checkups influenced by income level, representing the structural and socioeconomic conditions that shape preventive healthcare utilization. Together, these perspectives provide a more holistic understanding of how both personal and contextual factors contribute to preventive health outcomes in the United States.

Preventive healthcare plays a critical role in early disease detection and reducing long-term health costs . Regular checkups, screenings, and healthy lifestyle behaviors help identify potential health risks before they become serious, ultimately improving quality of life and reducing the financial burden on individuals and the healthcare system. Despite these benefits, engagement in preventive health practices varies widely among Americans, reflecting differences in awareness, motivation, and access to care. Research consistently shows that @wilcox2000 individuals who engage in consistent health-promoting behaviors, such as maintaining a balanced diet, exercising regularly, practicing stress management through meditation, and adhering to preventive medical routines, experience better long-term health outcomes. These include lower risks of chronic diseases such as hypertension, diabetes, and heart disease, as well as improved mental well-being and functional health throughout adulthood.

Previous research @rentfrow2013 shows that family income strongly influences access to medical care and preventive services. Adults with lower income levels face greater barriers to care, often forgoing medical visits or prescriptions due to cost, while even brief insurance disruptions reduce preventive services such as routine checkups. Building on this evidence, the present study investigates how socioeconoㅁmic factors—particularly annual household income—relate to both health effort and the time since one’s last routine checkup. By analyzing these relationships, this study seeks to highlight socioeconomic disparities in preventive healthcare among U.S. adults and contribute to a deeper understanding of the behavioral and structural determinants of health.

## Study Design and Data collection {#sec-design}

This study employs a cross-sectional observational design using data from the 2016–2017 American Health Values Survey, a nationally representative dataset of U.S. adults. The analysis focuses on respondents aged 18 and older who provided complete information on key variables, including self-reported effort in maintaining the health, blood pressure diagnosis, annual household income, and time since last routine medical checkup. Data were collected through structured survey questionnaires administered online and by phone, ensuring coverage across diverse demographic and socioeconomic groups. This design allows for examining associations between behavioral and socioeconomic factors and health outcomes without manipulating variables or establishing causality.

This study utilized a cross-sectional observational design drawing upon data from the 2016–2017 American Health Values Survey (AHVS), conducted by NORC at the University of Chicago with support from the Robert Wood Johnson Foundation. The AHVS is a nationally representative survey designed to assess the health beliefs, values, and behaviors of U.S. adults across diverse demographic and regional contexts. The analysis in this study was restricted to respondents aged 18 years and older who provided complete responses to key variables, including self-reported health effort, blood pressure diagnosis, annual household income, and time since last routine medical checkup.

Data were collected using a multimode survey design, incorporating online (CAWI), mail (self-administered paper questionnaires), and telephone interviews (CATI). This mixed-method approach enhanced accessibility for respondents with varying levels of internet access and literacy, ensuring more comprehensive coverage across urban and rural populations. The survey was available in both English and Spanish, and participants received small prepaid and contingent incentives to improve response rates

According to the AHVS and the related Sentinel Community Surveys collected detailed information on participants’ health values, personal health priorities, self-efficacy, access to care, and social determinants of health. The structured questionnaire captured both quantitative and categorical data, allowing for the examination of associations between behavioral, social, and economic factors and various health outcomes without any manipulation of variables. Because of its cross-sectional nature, this design allows for identifying correlations and group differences but does not infer causality.

## Methods

#### Data Preparation {#sec-dataprep-methods}

Data for this study were obtained from the 2016–2017 American Health Values Survey. The dataset was first cleaned and filtered to include only respondents aged 18 years and older to focus the analysis on adults. Observations with missing or incomplete responses for key analytical variables—specifically health effort, annual household income, and time since last routine medical checkup—were excluded to ensure data quality and consistency. This filtering process helped minimize bias that could result from incomplete information.

Before computing summary measures, several data-cleaning procedures were implemented. Responses coded as 77 (“Don’t know”) or 99 (“No answer”) were treated as invalid and replaced with missing values (NA) across relevant variables, particularly the health activity items (act1–act7) that contributed to the health effort measure. A new variable, healtheffort, was then created as the mean of act1 through act7 for each respondent, with missing values excluded from the calculation. This variable reflects an individual’s overall effort toward maintaining or improving their health.

The income and checkup variables underwent the same cleaning and recoding procedures. Responses coded as 77 (“Don’t know”) and 99 (“Refused”) were treated as missing values (NA). The income variable was categorized into eight ordered levels, ranging from less than \$15,000 to \$150,000 or more, while the checkup variable was recoded into five ordered groups, from within the past year to never. Data integrity checks were then performed to ensure that all invalid codes were removed and all transformations were accurately implemented. These procedures produced a reliable dataset suitable for statistical analysis and modeling.

#### Statistical Analysis {#sec-analysis-methods}

The statistical analysis for this study was guided by the analytical framework used in the American Health Values Survey (AHVS) and the Sentinel Community Health Values Surveys (SCHVS). The AHVS employed k-means clustering, a widely used unsupervised classification technique, to identify distinct groups of respondents based on their health values, beliefs, and behaviors. This method partitions the dataset into k mutually exclusive clusters by minimizing the within-cluster variance while maximizing the differences between cluster centroids.

Following the AHVS methodology, several solutions ranging from four to ten clusters were examined to determine the most meaningful group segmentation. The final cluster solution was selected based on statistical indicators such as the cubic clustering criterion and the Pseudo F statistic, ensuring that the identified typology best captured variation in respondents’ health attitudes and behaviors. Demographic and descriptive measures were then compared across clusters to confirm the face validity of the typology, ensuring that the observed groupings aligned with known differences in health, political, and socioeconomic characteristics among U.S. adults.

For the present analysis using the cleaned AHVS dataset, descriptive statistics were first computed to summarize respondent demographics and key health indicators (e.g., health effort, income level, and time since last routine checkup). Bivariate tests, including chi-square and correlation analyses, were used to assess relationships between categorical and continuous variables. Where applicable, multiple linear and logistic regression models were applied to examine associations between behavioral and socioeconomic predictors and health outcomes such as blood pressure diagnosis, while controlling for demographic covariates. All data cleaning, transformation, and analyses were performed using R (version 4.3.2). The tidyverse, dplyr, and ggplot2 packages were utilized for data manipulation and visualization, while stats functions supported regression modeling and significance testing. Statistical significance was set at p \< 0.05.

## Results {#sec-results}

```{r, fig-exercise, fig.cap=" Relationship between exercise effort and high blood pressure diagnosis.", fig.align='center', out.width='70%', include=TRUE, fig.show='hold', echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse); 
library(ggdist)
library(sjPlot); 
library(gtsummary)

load("~/Documents/GitHub/Draft/MATH_615_Draft/UShealth_clean.Rdata")

sjPlot::plot_xtab(
  grp=clean$bp_cat, 
  x=clean$exercise_cat,
  show.total = FALSE,
  margin="row",
  legend.title="") 

```

The bar chart illustrates the relationship between self-reported exercise effort and whether respondents reported a diagnosis of high blood pressure. Each bar shows the percentage of respondents in each exercise category (“Not at all” to “A great deal”) who reported having or not having high blood pressure.

As the graph shows, individuals who reported doing more exercise were less likely to have been diagnosed with high blood pressure. For example, among those who reported exercising “a great deal,” only 36.2% had high blood pressure, compared to 56.9% among those who did not exercise at all. This downward trend suggests that greater physical activity is associated with a lower likelihood of high blood pressure, supporting well-established evidence that regular exercise helps regulate blood pressure and improve cardiovascular health.

```{r}
load("~/Documents/GitHub/Draft/MATH_615_Draft/health_clean.Rdata")

sjPlot::plot_xtab(
  grp = clean$income_cat,
  x   = clean$checkup_cat,
  show.total = FALSE,
  margin = "row",
  legend.title = ""
)

```

The bar chart illustrates the relationship between household income and the recency of respondents’ last routine medical checkup. Individuals with higher income levels were more likely to have had a recent checkup, while those with lower incomes showed longer gaps or no checkups at all. This pattern indicates that higher-income adults have greater access to preventive healthcare, highlighting persistent socioeconomic disparities in medical care utilization.

## Discussion, Conclusion {#sec-discussino}

The results of this study reveal two key patterns linking behavioral and socioeconomic factors to preventive health outcomes among U.S. adults. First, individuals who reported putting greater effort into managing their personal health—through activities such as regular exercise, healthy eating, stress management, and proactive medical care—were significantly less likely to report a diagnosis of high blood pressure. This finding aligns with existing research demonstrating that consistent engagement in healthy behaviors helps regulate blood pressure, improve cardiovascular function, and reduce the risk of chronic disease. It reinforces the importance of self-care practices as an essential component of preventive healthcare, suggesting that promoting active health management could lead to improved long-term health outcomes and reduced healthcare costs.

Second, the analysis showed that household income and health insurance coverage were strong predictors of preventive healthcare utilization. Participants with higher income levels and stable insurance coverage were more likely to have received a recent routine medical checkup compared to those with lower income or no insurance. This pattern supports prior evidence that financial resources and access to care remain major determinants of preventive health engagement. The results highlight ongoing socioeconomic disparities in the U.S. healthcare system, where lower-income individuals may delay or forgo preventive visits due to cost, lack of coverage, or competing life demands.

Building on these findings, the analysis of routine checkup frequency by income level further illustrates this disparity. Individuals in lower income brackets showed smaller proportions of having had a checkup within the past year, whereas those in higher income brackets were more likely to have undergone a recent checkup. Notably, the “Never” category contained only respondents from middle-income groups, suggesting that preventive care avoidance may be tied to affordability and prioritization rather than complete neglect of healthcare. This gradient emphasizes how financial capability directly affects one’s likelihood of engaging in preventive medical visits—a pattern consistent with previous studies on healthcare access inequities.

Despite these meaningful insights, several limitations should be noted. The study’s cross-sectional design prevents establishing causal relationships between health effort, income, and health outcomes. The use of self-reported data may also introduce recall or social desirability bias, as individuals may overstate their healthy behaviors. Additionally, unmeasured factors such as education, neighborhood environment, or stress levels may influence both health behavior and outcomes.

Future research should explore these relationships using longitudinal data to better understand how personal health management and socioeconomic conditions interact over time. Including more detailed measures of lifestyle behaviors, health literacy, and psychological well-being could further clarify the mechanisms linking preventive efforts and health outcomes. Such work would strengthen the evidence base for policies and interventions aimed at improving access to preventive care and promoting equitable health behaviors across income groups.

In conclusion, the findings underscore that both behavioral commitment and socioeconomic stability are crucial to preventive healthcare engagement. Encouraging active self-management while addressing financial and structural barriers to care could help close the gap in preventive service utilization, ultimately leading to a healthier and more equitable population.

## References

::: {#refs}
:::

## Appendix

```{r, fig.show='hide'}
sjPlot::plot_xtab(grp=clean$bp_cat, x=clean$exercise_cat, 
                  show.total = FALSE, margin="row", legend.title="") 

```

```{r, fig.show='hide'}
sjPlot::plot_xtab(
  grp = clean$income_cat,
  x   = clean$checkup_cat,
  show.total = FALSE,
  margin = "row",
  legend.title = ""
)

```
